Changes in Faculty Composition and Stagnating State Support for Higher Education

### 2023-03-20 text/sections/00-abstract.tex
Finished major revision.  Needs more checks of writing style.

\noindent
State support for higher education per-student has stagnated at the same time as public universities have reduced their employment of tenure-track or tenured professors, and increased their employment of temporary contingent lecturers.
This analysis relates these changes in the faculty composition at public universities to the fall in per-student revenues from state funding for higher education.
A shift-share instrumental variables approach addresses endogeneity in decisions for state higher education funding by exploiting (1) yearly change in each state's funding for their public universities and (2) each institution's reliance on that support.
A decrease in state funding of 10\%, via a shock to state funding, decreases the number of assistant professors per student at a public university by 1.4\% and full professors by 1.2\%, yet increases the number of lecturers per student by 4.4\%.
Local projection estimates show that these effects linger for up to three years after the initial funding shock.
Analysis of all the professors at Illinois public universities 2011-2021 shows that incumbent professors are not affected by the changes in state support, while tenured-track hiring counts fall over this time period, showing that these changes in faculty composition arose by reduced hiring of tenure-track and tenured professors at public universities.
These results show the long-term effects of stagnating state support for higher education, and raises questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

\noindent
\textbf{Key-words:}
State and Local Budget and Expenditures,
Higher Education,
Public Sector Labour Markets

\noindent
\textbf{JEL Codes:} H72, I23, J45

>>>>

\noindent
State support for higher education per-student has stagnated while universities have substituted towards contingent lecturers and away from tenure-track/tenured professors.
This analysis relates these changes in the faculty composition at public universities to the fall in state funding per student for higher education, using a shift-share approach to identify changes in a state funding among public universities.
A decrease in state funding of 10\%, via a shock to state funding, decreases the number of assistant professors per student at a public university by 1.4\% and full professors by 1.2\%, yet increases the number of lecturers per student by 4.4\%.
Local projection estimates show that these effects linger for up to three years after the initial funding shock.
Analysis of all the professors at Illinois public universities 2011-2021 shows that incumbent professors are not affected by the changes in state funding, and suggestive evidence shows that these changes arose by impacting the hiring of new professors at public universities impacted by the funding shocks.
These results show the long-term effects of stagnating state support for higher education, and raise questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

\vfill
\noindent
\textbf{Key-words:}
State and Local Budget and Expenditures,
Higher Education,
Public Sector Labour Markets

\vspace{0.1cm}
\noindent
\textbf{JEL Codes:} H72, I23, J45

### 2023-03-20 text/sections/01-introduction.tex
Finished major revision.  Needs more checks of writing style.

Negative changes in state funding affects faculty composition at public universities, away from tenured and tenure-track professors towards non-tenure track lecturer positions.
A reduction of 10\% in state funding, via a shock to state appropriations, leads to a fall in 1.4\% in the number of assistant professors per student within a university, 1.2\% for full (tenured) professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for three to four years after the initial funding shock.
Over the same time period, state funding fell by around 35\%, while the count of professors per student fell by 9\%, so that these results show that falls in state funding explain around a third of the observed shift away from tenure-track and tenure professors towards contingent faculty.

Additionally, I use individual level data on professors at Illinois public universities over 2010-2021 to investigate whether the shocks to state funding affected individual professors.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet, the hiring rate of tenure-track professors declined over this time period for all public universities.
This shows that changes in faculty composition came about by hiring reductions of new professors at public universities.

Absent this analysis, it is not immediately clear why stagnating state support and changes in public university faculty composition would be causally linked.
Over the 21st Century there have been drastic changes to college selectivity, where the top universities have become more selective while the average university less selective \citep{hoxby2009changing}.
While the average public university is becoming less selective, it is possible that their most productive research-focused faculty more often move to more selective and prestigious (and often private) institutions, so that over time public universities (on average) substitute towards contingent lecturers. 
This is one way in which the trends may be concurrent, but not causal.
This analysis shows that stagnating in state funding for public universities \textit{cause} the changes in faculty composition.
The state funding channel causes about 40\% of the observed substitution towards contingent lecturers, so that it is possible that other systematic changes in higher education (such as changing selectivity) may explain the rest of the variation.

@@

The number of professors, both in absolute terms and relative to number of students, may be particularly important for quality of instruction and thus educational outcomes.
Small class sizes class size at high school lead to increases in test scores \citep{angrist1999using}, and UK university students perform worse academically in particularly large classes \citep{bandiera2010heterogeneous}.
Regarding faculty composition, the literature is not decided on whether contingent lecturers lead to better or worse education outcomes relative to tenured faculty.
Adjunct professors are shown to better undergraduate teachers at highly selective universities \citep{bettinger2010does,figlio2015tenure}, yet universities who use a greater share of lecturers have lower graduation rates \citep{ehrenberg2005tenured}.
Notably, students taught by part-time adjuncts lecturers have worse student outcomes than full-time professors at a less-selective university \citep{zhu2021limited}, but this difference disappears when individual lecturers gain a full-time teaching position at the university. 
My approach focuses on count of professors per student at universities, and considers faculty count and composition change as possible mechanisms for how stagnating state support for higher education negatively impacts student outcomes \citep{NBERw23736,NBERw27885}.

>>>>

Negative changes in state funding affects faculty composition at public universities, away from tenured and tenure-track professors towards non-tenure track lecturer positions.
A reduction of 10\% in state funding, via a shock to state appropriations, leads to a fall in 1.4\% in the number of assistant professors per student within a university, 1.2\% for full (tenured) professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for three to four years after the initial funding shock.
Over the same time period, state funding fell by around 35\%, while the count of professors per student fell by 9\%, so that these results show that falls in state funding explain around a third of the observed shift away from tenure-track and tenure professors towards contingent faculty.
Additionally, I use individual level data on professors at Illinois public universities over 2010-2021 to investigate whether the shocks to state funding affected individual professors.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet, the hiring rate of tenure-track professors declined over this time period for all public universities.
This shows that changes in faculty composition came about by hiring reductions of new professors at public universities.

Absent this analysis, it is not immediately clear why stagnating state support and changes in public university faculty composition would be causally linked.
Over the 21st Century there have been drastic changes to college selectivity, where the top universities have become more selective while the average university less selective \citep{hoxby2009changing}.
While the average public university is becoming less selective, it is possible that their most productive or research-focused faculty more often move to more selective and prestigious (and often private) institutions, so that over time public universities (on average) substitute towards contingent lecturers. 
This is one way in which the trends may be concurrent, but not causal.
This analysis shows that stagnating in state funding for public universities cause the changes in faculty composition, explaining about 40\% of the observed substitution towards contingent lecturers, so that it is possible that other systematic changes in higher education (such as changing selectivity) may explain the rest of the variation.

@@

The number of professors, both in absolute terms and relative to number of students, may be particularly important for quality of instruction and thus educational outcomes.
Small class sizes in secondary schools lead to increases in test scores \citep{angrist1999using}, and UK university students perform worse academically in particularly large classes \citep{bandiera2010heterogeneous}.
Regarding faculty composition, the literature is not decided on whether contingent lecturers lead to better or worse education outcomes relative to tenured faculty.
Adjunct professors are shown to better undergraduate teachers at highly selective universities \citep{bettinger2010does,figlio2015tenure}, yet universities who use a greater share of lecturers have lower graduation rates \citep{ehrenberg2005tenured}.
Notably, students taught by part-time adjuncts lecturers have worse student outcomes than full-time professors at a less-selective university \citep{zhu2021limited}, but this difference disappears when individual lecturers gain a full-time teaching position at the university. 
My approach focuses on count of professors per student at universities, and considers faculty count and composition change as possible mechanisms for how stagnating state support for higher education negatively impacts student outcomes \citep{NBERw23736,NBERw27885}.
